---
title: "R packages structure research projects and enhance reproducibility"
author:
  - Frederick J. Boehm:
      email: frederick.boehm@gmail.com  
      institute: [UofM]
      correspondence: true
  - Bret M. Hanlon:
      email: bmhanlon@wisc.edu
      institute: [UofW]
      correspondence: false
institute:
  - UofM: University of Michigan
  - UofW: University of Wisconsin-Madison
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  We present a case for using R packages as the core structure for research projects. We review the ways in which R packages may be used for teaching research skills in data science and statistics. Building on existing software tools, we present reproducible R code that enables users to construct their own packages for use in data analysis and research. The R package structure has much to offer teachers and students of data science and statistics. Below, we review the standard structure of R packages. We then elaborate this structure to include a subdirectory for additional data analysis resources. Our audience is instructors of courses for upper level undergraduate students. We assume that these undergraduate students have modest familiarity with R computing, presumably developed in data science and statistics courses and independent research projects, so we don't discuss approaches to teaching basic R skills. 
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

## R package structure

The structure of an R package consists of a collection of specifically named directories and files. Every R package must have:

TABLE 1: List of required files and directories for an R package.
(Not sure that I can name these off the top of my head, but it's easy to look them up). Files: DESCRIPTION, NAMESPACE, etc
Directories: R, man, ...

* Need to explain contents and structure of each item in TABLE 1. 


## How to construct an R package

* usethis R package; devtools R package; rrtools R package on github
* Provide R exact code here
* Rstudio IDE shortcuts
* Github actions (added to package via calls to usethis functions)

We demonstrate one way to construct an R package from scratch. We do this from within an R session. We also discuss how users of Rstudio IDE software can quickly and efficiently start a new R package. 

What subdirectories to add to a "mypackage/analysis" directory:
Rscript: .R files that 
Rmd: Rmarkdown files and their outputs
data: inputs of any format
results: tables, figures, intermediate rds files?



## Teaching others how to construct an R package

* Existing resources:
    * Check Greg Wilson's book Teaching Technology Together
    * Data Carpentry resources?? I'm unaware of R package materials from The Carpentries
    * Rstudio's online refs

* Karl Broman's tools4rr page: https://kbroman.org/Tools4RR/
* 



## Assessment of R packages for data analysis projects

* Include a detailed rubric in our paper
* Scaffolding a long-term project with intermediate deadlines

<!-- Here's some example analysis code: -->

```{r get-data, eval = FALSE}
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r demo-plot, fig.cap="A plot of random numbers"}
plot(rnorm(10))
```

Figure \@ref(fig:demo-plot) shows how we can have a caption and cross-reference for a plot

```{r demo-inline-code}
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence. 

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References 

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
